---
title: "A Pragmatic Approach to Census Analysis:\n Tidycensus and R"
author: "Jamaal Green"
date: "September 19, 2017"
geometry: margin=2cm
output: 
  beamer_presentation:
    incremental: true
    theme:  "Boadilla"
    colortheme: "beetle"
    slide_level: 2
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, tidy = FALSE)
```

#Introduction
```{r, message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
```

##About Me

- PhD Candidate in Urban Studies and Planning
- My dissertation is examining industrial zoning and labor market change
- I use a pretty wide array of census products for work (ACS, PUMS, LEHD)

##A Pragmatic Approach

"Let the question guide your method"

Likewise...

Let your problems guide your tools

##What's your workflow normally resemble?

The steps many social data analysts and GIS user have to make:

- Tabular Data collection/import (factfinder)
- Spatial Data collection(Tigerline files, anyone?)
- Tabular data cleaning, munging, and joins
- Table to spatial data joins (we've all done this in Arc with moderate success)
- If making maps...spatial processing (clips, intersections, spatial joins)
- Other visualizations and report writing

##This Works But It Could be Better...

This workflow is effective but suffers from:

1.  Massive number of intermediate outputs 
2.  Jumps among any number of different applications making confusion likely
3.  Can easily become disorganized if data/project management isn't specified beforehand

#Enter R

##What is R?

- A powerful language
- Application
- "Do it all" workbench

##But why R?

- It's free
- It's fast
- It's data type agnostic (read any variety of text files, .shp. GEOJSON, GEOTIFF)
- Massive number of packages for statistical or spatial analysis and visualization
- Many things that are hard or slow in other applications (table joins in Arc, anyone?) are fast in R 
- Large, helpful online community and growing variety of books/guides/courses
- IT'S FREE

##But why should I?

Has the following ever happened to you?

- Need to download multiple variables over multiple years and you get a data folder filled with ambiguously named tables that you end up deleting anyway?
- Had to change your geography of interest on short notice and then go through the time consuming process of redownloading and processing?
- Attempted to rename a column in ArcMap (yes, I know this is now available in ArcPro)?


##Let's Be Pragmatic

These recurring challenges can be better addressed (saving yourself precious time) by learning a little bit of R 

#Enter the tidyverse and tidycensus...A Better Way

##tidyverse

"An opinionated collection of R packages for data science"

A set of packages to handle common "data science" tasks with consistent behavior and language. A more accessible way to do data science in R for all steps of a project from data import/cleaning to visualization and modeling


##Tidycensus...one stop shop for ACS data

- R package authored by Prof. Kyle Walker at TCU to make gathering and visualizing census data easier
- The package uses the census API to call ACS and decennial data as well as ACS Data Profile tables
- Data is returned in either wide or long format and there is an option to join the data to its appropriate Tigerline geometry

##A quick example...

Our assignment: Get latest 5 year MHI for Multnomah County at Tract Level and graph the results (as an added bonus, and in the interest of transparency, let's include MOEs)

##Query ACS

```{r, eval=FALSE, message=FALSE, warning=FALSE}
if(!require(pacman)){install.packages("pacman"); 
  library(pacman)}
p_load(ggplot2, tidycensus, dplyr)

acs_key <- Sys.getenv("CENSUS_API_KEY")

#Enter the variables and geographies below
census_title <- c("Median Household Income by County:\n 
Coefficient of Variation")
census_var <- c("B19013_001E")
census_geog <- c("county")
census_state <- c("or")


acs_data <- get_acs(geography = census_geog, variables = 
census_var, state = census_state, output = "wide")

```

##Little Bit of Processing

```{r, eval=FALSE, message=FALSE, warning=FALSE}

#Make more readable column names
acs_data <- acs_data %>% 
  rename(MHI_est = B19013_001E , 
   MHI_moe = B19013_001M)

#Calculate the SE, CV for future reference
acs_data <- acs_data %>% 
  mutate(se = MHI_moe/1.645, 
         cv = (se/MHI_est)*100)
```


##Finally...let's plot

```{r, eval=FALSE, message=FALSE, warning=FALSE}


#Plot Percentages with Derived MOE
acs_plot <- acs_data %>%
  ggplot(aes(x = MHI_est, 
  y = reorder(NAME, MHI_est))) +
  geom_point(color = "black", size = 2) +
  geom_errorbarh(aes(xmin = MHI_est - MHI_moe, 
                     xmax = MHI_est + MHI_moe )) +
  labs(title = paste(census_title),
       subtitle = 
      paste0("Oregon 2011-2015 American Community Survey"),
       x = "Median Household Income") + 
  scale_x_continuous(labels = scales::dollar) + 
  theme_minimal() +
  theme(panel.grid.minor.x  = element_blank(),
        panel.grid.major.x = element_blank())

plot(acs_plot)

```

##Our Output

```{r, echo=FALSE, message=FALSE, warning=FALSE}
if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(ggplot2, tidycensus, dplyr)


#key <- census_api_key("c67f1b3134300374d51a55c543649f843fb7d2b3",install = TRUE)
acs_key <- Sys.getenv("CENSUS_API_KEY")

#Enter the variables and geographies below
census_title <- c("Median Household Income by County:\n Coefficient of Variation")
census_var <- c("B19013_001E")
census_geog <- c("county")
census_state <- c("or")


acs_data <- get_acs(geography = census_geog, variables = census_var, state = census_state, 
                    output = "wide")

#Make more readable column names
acs_data <- acs_data %>% rename(MHI_est = B19013_001E , MHI_moe = B19013_001M)

#Calculate the SE, CV for future reference
acs_data <- acs_data %>% mutate(se = MHI_moe/1.645, cv = (se/MHI_est)*100)



#Plot Percentages with Derived MOE
acs_plot <- acs_data %>%
  ggplot(aes(x = MHI_est, y = reorder(NAME, MHI_est))) +
  geom_point(color = "black", size = 2) +
  geom_errorbarh(aes(xmin = MHI_est - MHI_moe, xmax = MHI_est + MHI_moe )) +
  labs(title = paste(census_title),
       subtitle = paste0("Oregon 2011-2015 American Community Survey"),
       y = "",
       x = "Median Household Income") + 
  scale_x_continuous(labels = scales::dollar) + theme_minimal() +
  theme(panel.grid.minor.x  = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.background = element_rect(fill = "white"))

print(acs_plot)
```

#Mapping It Out

Your Assignment: map median household income at the census tract level for the PDX MSA OR counties and for the counties at the state level

##R as a GIS- tigris and sf

tigris- a package that will download tigerline shapefiles

simple features- uses well known text to signify geometry allowing for spatial objects to be treated as dataframes



##Tract Processing tidyverse style

```{r, eval=FALSE, message=FALSE, warning=FALSE}
if(!require(pacman)){install.packages("pacman"); 
  library(pacman)}
p_load(sf, tigris, viridis, ggthemes, ggplot2, 
       tidycensus, stringr, dplyr)
options(tigris_class = "sf", tigris_use_cache = TRUE)

acs_key <- Sys.getenv("CENSUS_API_KEY")

mhi_tables <- c("B19013_001")

#download tracts and county, get the tracts for PDX 
#Metro counties and counties for the state

mhi_tract <- get_acs(geography = "tract", 
                     variables = mhi_tables, 
                     state = "OR", 
                     geometry = TRUE)

mhi_tract <- mhi_tract %>% 
  mutate(CountyFIPS = str_sub(GEOID, 1, 5))

metro_counties <- c("41051", "41005", "41009", "41067", "41071") 
metro_tract <- mhi_tract %>% 
  filter(CountyFIPS %in% metro_counties)

or_mhi_county <- get_acs(geography ="county", variables = mhi_tables, state = "OR", geometry = TRUE)
```

##Our Tract Map Set Up

```{r, eval=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot() + 
  geom_sf(data = metro_tract, aes(fill = estimate)) +
  coord_sf(datum = NA) + 
  theme(plot.title = element_text(size = 16, 
                    face = "bold", margin = margin(b=10))) +
  theme(plot.subtitle = element_text(size = 14, 
                      margin = margin(b = -20))) +
  theme(plot.caption = element_text(size = 9, 
                    margin = margin(t = -15), hjust = 0)) +
  scale_fill_viridis(labels = scales::dollar, 
                     name = "MHI Estimate") +
  labs(caption = "Source: US Census Bureau ACS (2011-2015)",
        title = "Median Household Income for PDX Metro\n at the census tract level",
        subtitle = "An R 'sf' Example") + theme_minimal()


```

##Our Tract Output

```{r, echo=FALSE, message=TRUE, warning=FALSE}
if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(sf, tigris, viridis, ggthemes, ggplot2, tidycensus, stringr, dplyr)
options(tigris_class = "sf", tigris_use_cache = TRUE)

#stretch example is MHI at census tract level for METRO counties
#gonna collect census tracts with geography, sum up to counties for graph

#key <- census_api_key("c67f1b3134300374d51a55c543649f843fb7d2b3",install = TRUE)
acs_key <- Sys.getenv("CENSUS_API_KEY")

vartabs <- load_variables(year = 2015, dataset = "acs5")

mhi_tables <- c("B19013_001")

#download tracts and county, get the tracts for PDX Metro counties and counties for the state

mhi_tract <- get_acs(geography = "tract", variables = mhi_tables, state = "OR", geometry = TRUE)
mhi_tract <- mhi_tract %>% mutate(CountyFIPS = str_sub(GEOID, 1, 5))

metro_counties <- c("41051", "41005", "41009", "41067", "41071") 
metro_tract <- mhi_tract %>% filter(CountyFIPS %in% metro_counties)

or_mhi_county <- get_acs(geography ="county", variables = mhi_tables, state = "OR", geometry = TRUE)

#Metro tract map

p1 <- ggplot() + geom_sf(data = metro_tract, aes(fill = estimate)) +
  coord_sf(datum = NA) + 
  theme(plot.title = element_text(size = 16, face = "bold", margin = margin(b=10))) +
  theme(plot.subtitle = element_text(size = 14, margin = margin(b = -20))) +
  theme(plot.caption = element_text(size = 9, margin = margin(t = -15), hjust = 0)) +
  scale_fill_viridis(labels = scales::dollar, name = "MHI Estimate") +
  labs( caption = "Source: US Census Bureau ACS (2011-2015)",
        title = "Median Household Income for PDX Metro\n at the census tract level",
        subtitle = "An R 'sf' Example") + theme_minimal()

plot(p1)
```

##Our County Output

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#state county map

p2 <- ggplot() + 
  geom_sf(data = or_mhi_county, aes(fill = estimate)) +
  coord_sf(datum = NA) +
  theme(plot.title = element_text(size = 14, 
                    margin = margin(b = -20))) +
  theme(plot.subtitle = element_text(size = 9, 
                      margin = margin(t = -15), hjust = 0)) +
  theme(plot.caption = element_text(size = 9, 
                    margin = margin(t = -15), hjust = 0)) +
  scale_fill_viridis(labels = scales::dollar, 
                     name = "MHI Estimate") +
  labs(caption = "Source: US Census Bureau ACS (2011-2015)",
       title = "Median Household Income for Oregon\n at the county level",
       subtitle = "An R 'sf' Example") + theme_minimal()

plot(p2)
```

#Let's Stretch A Bit

##Your Assignment

Create a social vulnerability index for the PDX Metro area using % in poverty, non-White %, % under 5 years of age, and % over 64 years of age and map it

##Inital Collection and Processing

```{r, eval=FALSE, message=FALSE, warning=FALSE}
#name the tables we need
vul_vars <- c("B17001_001", "B17001_002",  "B02001_001", 
              "B02001_002", "B01001_003", "B01001_020",
              "B01001_021", "B01001_022", "B01001_023", 
              "B01001_024", "B01001_025","B01001_027", 
              "B01001_044", "B01001_045", "B01001_046", 
              "B01001_047", "B01001_048", "B01001_049")
              
#grab the data for Oregon

vul_acs <- 
  get_acs(geography = "tract", variables = vul_vars, 
  state = "OR", output = "wide")

vul_acs <- vul_acs %>% 
  mutate(CountyFIPS = str_sub(GEOID, 1, 5))
```

##Clean Up Table and Calculate Percentages

```{r, eval=FALSE, message=FALSE, warning=FALSE}
vul2 <- vul_acs %>% 
  mutate(PovShare = B17001_002E/B17001_001E,
        NonWhite = (B02001_001E - B02001_002E)/B02001_001E,
        Under5 = (B01001_003E + B01001_027E)/B02001_001E,
        Older64Male = B01001_020E + B01001_021E + 
        B01001_022E +B01001_023E + B01001_024E +
        B01001_025E, 
        Older64Female = B01001_044E + 
        B01001_045E + B01001_046E + B01001_047E + 
        B01001_048E + B01001_049E, 
        Older64 = 
        (Older64Male + Older64Female)/B02001_001E) %>% 
  select(NAME, GEOID, CountyFIPS, PovShare, 
         NonWhite, Under5, Older64)
  
```

##Get Index Values

```{r , eval=FALSE, message=FALSE, warning=FALSE, width = 45}
vul2 <- vul2 %>% 
  mutate(
  z_Pov = (PovShare - mean(PovShare, na.rm = TRUE))/
           sd(PovShare, na.rm = TRUE),
  z_NonWhite = (NonWhite - mean(NonWhite, na.rm = TRUE))/
           sd(NonWhite, na.rm = TRUE),
  z_Under5 = (Under5 - mean(Under5, na.rm = TRUE))/
           sd(Under5, na.rm = TRUE),
  z_Older64 = (Older64 - mean(Older64, na.rm = TRUE))/
           sd(Older64, na.rm = TRUE))

vul2 <- vul2 %>% 
  mutate(VulIndex = (z_Pov + z_NonWhite + z_Under5 + 
                      z_Older64)/4) %>% 
  select(GEOID, CountyFIPS, z_Pov, z_NonWhite, z_Under5, 
         z_Older64, VulIndex)

```

##Subset & Join to Geometry

```{r, eval=FALSE, message=FALSE, warning=FALSE}

metro_counties <- c("41051", "41005", "41009", 
                    "41067", "41071") 

vul2 <- vul2 %>% 
  filter(CountyFIPS %in% metro_counties)

or_tracts <- tracts(state = "OR")

metro_vul <- inner_join(vul2, or_tracts, 
                        by = c("GEOID" = "GEOID")) %>% 
  select(1:7, geometry) %>% st_as_sf()
  
```

##And...voila

```{r, echo=FALSE, message=FALSE, warning=FALSE}

options(tigris_class = "sf")

vul_vars <- c("B17001_001", "B17001_002",  "B02001_001", "B02001_002", "B01001_003", 
              "B01001_020","B01001_021", "B01001_022", "B01001_023", "B01001_024", 
              "B01001_025","B01001_027", "B01001_044", "B01001_045", "B01001_046", 
              "B01001_047", "B01001_048", "B01001_049")

vul_acs <- get_acs(geography = "tract", variables = vul_vars, state = "OR", output = "wide")
vul_acs <- vul_acs %>% mutate(CountyFIPS = str_sub(GEOID, 1, 5))

vul2 <- vul_acs %>% mutate(PovShare = B17001_002E/B17001_001E,
                          NonWhite = (B02001_001E - B02001_002E)/B02001_001E,
                           Under5 = (B01001_003E + B01001_027E)/B02001_001E,
                           Older64Male = B01001_020E + B01001_021E + B01001_022E +
                             B01001_023E + B01001_024E +B01001_025E, 
                           Older64Female = B01001_044E + B01001_045E + B01001_046E +
                             B01001_047E + B01001_048E + B01001_049E, 
                           Older64 = (Older64Male + Older64Female)/B02001_001E) %>% 
  select(NAME, GEOID, CountyFIPS, PovShare, NonWhite, Under5, Older64)

vul2 <- vul2 %>% 
  mutate(z_Pov = (PovShare - mean(PovShare, na.rm = TRUE))/sd(PovShare, na.rm = TRUE),
         z_NonWhite = (NonWhite - mean(NonWhite, na.rm = TRUE))/sd(NonWhite, na.rm = TRUE),
         z_Under5 = (Under5 - mean(Under5, na.rm = TRUE))/sd(Under5, na.rm = TRUE),
         z_Older64 = (Older64 - mean(Older64, na.rm = TRUE))/sd(Older64, na.rm = TRUE))

vul2 <- vul2 %>% 
  mutate(VulIndex = (z_Pov + z_NonWhite + z_Under5 + z_Older64)/4) %>% 
  select(GEOID, CountyFIPS, z_Pov, z_NonWhite, z_Under5, z_Older64, VulIndex)

metro_counties <- c("41051", "41005", "41009", "41067", "41071") 

vul2 <- vul2 %>% filter(CountyFIPS %in% metro_counties)

or_tracts <- tracts(state = "OR")

metro_vul <- inner_join(vul2, or_tracts, by = c("GEOID" = "GEOID")) %>% 
  select(1:7, geometry) %>% st_as_sf()

p4 <- ggplot() + geom_sf(data = metro_vul, aes(fill = VulIndex)) +
  coord_sf(datum = NA) +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 14),
        plot.caption = element_text(size = 9)) +
  scale_fill_viridis(name = "Social Vulnerability Measure") +
  labs(title = "Social Vulnerability for Metro oregon",
       subtitle = "An R 'sf' example",
       caption = "Source: US Census Bureau ACS (2011-2015)") +
  theme_minimal()

plot(p4)

```

#Wrapping Up

##Let the Problem Guide The Tool

It's not necessary to do everything in R, but we can do a lot of things faster and easier in R

##A tidy approach...

The tidyverse, and packages connected to it, make R more approachable than ever 

##Some Additional Resources

> - [\textcolor{blue}{"R For Data Science" by Wickham and Grolemund}](http://r4ds.had.co.nz/)
> - [\textcolor{blue}{DataCamp}](https://www.datacamp.com/home)
> - [\textcolor{blue}{StackOverflow}](https://stackoverflow.com/questions/tagged/r)
> - [\textcolor{blue}{Presentation Link}](https://github.com/jamgreen/acs_presentation)