---
title: 'Census, LEHD and Planning: An R Approach'
author: "Jamaal Green"
institute: "Portland State University"
date: "12/10/2018"
output: 
  beamer_presentation:
    theme: "Montpellier"
    colortheme: "beetle"
    fonttheme: "structurebold"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# Introduction


## Good Morning

- Doctoral candidate here in TSUSP, planning on defending this spring
- My dissertation focuses on issues surrounding industrial land use policy
- My GRA work deals with a mix of mapping, some spatial analysis, census data wrangling and some modeling

## Goals for today

- Discuss *tidycensus* and my day-to-day workflows
- Introducing *lehdr*
- Dissertating with R

# Tidycensus and the Day-to-Day

## Tidycensus- Game Changer

- *tidycensus* is an R packaged authored and maintained by Prof. Kyle Walker at TCU
- a wrapper of the censusapi package and simplifies calls to grab ACS data, in particular
- radically simplifies and streamlines traditional census workflows, particularly with GIS and gives us data in a "tidy" format

## Tigris

- a package for grabbing Tiger/Line shapefiles
- also authored by Kyle Walker
- *tidycensus* depends on tigris bringing tabular and spatial data calls in one


## My old workflow

- Tabular Data collection/import (FactFinder or DataFerret for ACS/Decennial)
- Spatial Data collection(Tigerline files)
- Tabular data cleaning and munging in excel 
- Table to spatial data joins in Arc (successful maybe 40% of the time on first attempt) 
- If making maps...spatial processing (clips, intersections, spatial joins)
- Other visualizations and report writing (excel, word, and ppt)

##  Can tidycensus do all that?!

>-  No
>-  But tidycensus, in particular, directly solves aspects of the first three areas
>-  In combination with additional packages from the tidyverse (particularly *dplyr* and *ggplot*) and the spatial package *sf* all data processing can be done with R

## An Example

Our assignment: Get latest 5 year MHI for Multnomah County at Tract Level and graph the results (as an added bonus, and in the interest of transparency, let's include CVs)

```
install.packages("pacman"); library(pacman)
p_load(ggplot2, tidycensus, dplyr)

acs_key <- Sys.getenv("CENSUS_API_KEY")

#Enter the variables and geographies below
census_title <- c("Median Household Income by County:\n 
Margin of Error")
census_var <- c("B19013_001E")
census_geog <- c("county")
census_state <- c("or")
```

## The tidycensus call

```
acs_data <- get_acs(geography = census_geog, variables = 
census_var, state = census_state, output = "wide")

#Make more readable column names
acs_data <- acs_data %>% 
rename(MHI_est = B19013_001E , MHI_moe = B19013_001M)

#Calculate the SE, CV for future reference
acs_data <- acs_data %>% 
mutate(se = MHI_moe/1.645, cv = (se/MHI_est)*100)

```

## Finally, to the plot...

```
#Plot Percentages with Derived MOE
acs_plot <- acs_data %>%
  ggplot(aes(x = MHI_est, y = reorder(NAME, MHI_est))) +
  geom_point(color = "black", size = 2) +
  geom_errorbarh(aes(xmin = MHI_est - MHI_moe, 
  xmax = MHI_est + MHI_moe )) +
  labs(title = paste(census_title),
       subtitle = paste0("Oregon 2011-2015 American 
       Community Survey"),
       y = "",
       x = "Median Household Income") + 
  scale_x_continuous(labels = scales::dollar) + 
  theme_minimal() +
  theme(panel.grid.minor.x  = element_blank(),
        panel.grid.major.x = element_blank())

plot(acs_plot)
```


## Sample Output 

```{r echo=FALSE, message=FALSE, warning=FALSE}
if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(ggplot2, tidycensus, dplyr)


#key <- census_api_key("c67f1b3134300374d51a55c543649f843fb7d2b3",install = TRUE)
acs_key <- Sys.getenv("CENSUS_API_KEY")

#Enter the variables and geographies below
census_title <- c("Median Household Income by County:\n Margin of Error")
census_var <- c("B19013_001E")
census_geog <- c("county")
census_state <- c("or")


acs_data <- get_acs(geography = census_geog, variables = census_var, state = census_state, 
                    output = "wide")

#Make more readable column names
acs_data <- acs_data %>% rename(MHI_est = B19013_001E , MHI_moe = B19013_001M)

#Calculate the SE, CV for future reference
acs_data <- acs_data %>% mutate(se = MHI_moe/1.645, cv = (se/MHI_est)*100)



#Plot Percentages with Derived MOE
acs_plot <- acs_data %>%
  ggplot(aes(x = MHI_est, y = reorder(NAME, MHI_est))) +
  geom_point(color = "black", size = 2) +
  geom_errorbarh(aes(xmin = MHI_est - MHI_moe, xmax = MHI_est + MHI_moe )) +
  labs(title = paste(census_title),
       subtitle = paste0("Oregon 2011-2015 American Community Survey"),
       y = "",
       x = "Median Household Income") + 
  scale_x_continuous(labels = scales::dollar) + theme_minimal() +
  theme(panel.grid.minor.x  = element_blank(),
        panel.grid.major.x = element_blank())

print(acs_plot)
```

## Extending the package

 Let's make it spatial...going to map MHI at tract level for the PDX MSA
 
## Setting it up
```
mhi_tables <- c("B19013_001")

mhi_tract <- get_acs(geography = "tract", 
                     variables = mhi_tables, 
                     state = "OR", 
                     geometry = TRUE)

mhi_tract <- mhi_tract %>% 
  mutate(CountyFIPS = str_sub(GEOID, 1, 5))

metro_counties <- c("41051", "41005", "41009", "41067", 
"41071") 

metro_tract <- mhi_tract %>% 
  filter(CountyFIPS %in% metro_counties)
  
```
\end{small}

## Finalizing plot with *sf* and *ggplot*

```
p1 <- ggplot() + 
  geom_sf(data = metro_tract, aes(fill = estimate)) +
  coord_sf(datum = NA) + 
  theme(plot.title = element_text(size = 16, 
                    face = "bold", margin = margin(b=10))) +
  theme(plot.subtitle = element_text(size = 14, 
                      margin = margin(b = -20))) +
  theme(plot.caption = element_text(size = 9, 
                    margin = margin(t = -15), hjust = 0)) +
  scale_fill_viridis(labels = scales::dollar, 
                     name = "MHI Estimate") +
  labs(caption = "Source: US Census Bureau ACS (2011-2015)",
        title = "Median Household Income for PDX Metro\n 
        at the census tract level",
        subtitle = "An R 'sf' Example") + theme_minimal()


```

## Our result

```{r echo=FALSE, message=FALSE, warning=FALSE}

if(!require(pacman)){install.packages("pacman"); 
  library(pacman)}
p_load(sf, tigris, viridis, ggthemes, ggplot2, 
       tidycensus, stringr, dplyr)
options(tigris_class = "sf", tigris_use_cache = TRUE)

mhi_tables <- c("B19013_001")


mhi_tract <- get_acs(geography = "tract", 
                     variables = mhi_tables, 
                     state = "OR", 
                     geometry = TRUE)

mhi_tract <- mhi_tract %>% 
  mutate(CountyFIPS = str_sub(GEOID, 1, 5))

metro_counties <- c("41051", "41005", "41009", "41067", "41071") 
metro_tract <- mhi_tract %>% 
  filter(CountyFIPS %in% metro_counties)

p1 <- ggplot() + 
  geom_sf(data = metro_tract, aes(fill = estimate)) +
  coord_sf(datum = NA) + 
  theme(plot.title = element_text(size = 16, 
                    face = "bold", margin = margin(b=10))) +
  theme(plot.subtitle = element_text(size = 14, 
                      margin = margin(b = -20))) +
  theme(plot.caption = element_text(size = 9, 
                    margin = margin(t = -15), hjust = 0)) +
  scale_fill_viridis(labels = scales::dollar, 
                     name = "MHI Estimate") +
  labs(caption = "Source: US Census Bureau ACS (2011-2015)",
        title = "Median Household Income for PDX Metro\n at the census tract level",
        subtitle = "An R 'sf' Example") + theme_minimal()

plot(p1)

```

## From there

- From here I would use RMarkdown for report writing, back up with version control in github and go from there
- While a simplistic example, *tidycensus* along with other packages have freed my daily workflow from the time suck of going back and forth from FactFinder or DataFerret for common census datasets as well as switching from desktop GIS to excel

# *lehdr*- Freeing LODES data

## *lehdr*

- Package I designed with Profs. Liming Wang here at PSU and Dillon Mahmoudi at UMBC
- Comes out of my dissertation work to more efficienctly grad LEHD LODES data
- Hope to release on CRAN in 2019

## LEHD LODES

- Longitudinal Employer-Household Dynamics Dataset, Longitudinal Origin-Destination Employment Series
- A mouthful, but LODES gives us the finest geographic detail on employment (only 2 digit NAICS) for the country, as well as origin-destination tables
- You may be more familiar with [OnTheMap](https://onthemap.ces.census.gov/)

## Inspiration

- Dissertation requires looking at change in mfg employment between 2009-2015 at block group level
- Needed to grab multiple states for multiple time periods without using the ftp site manually
- Built up on from an unwiedly to relatively neat package designed to play nicely with tidy tools

## Manufacturing in Multnomah County

- My dissertation looks at urban mfg and land use
- Can we take a look at how the geography of mfg employment has changed over time in Multnomah County (2005-2015)

## The Set Up

```
if(!require(pacman)){install.packages("pacman"); 
library(pacman)}
p_load(devtools, dplyr, tigris, sf, stringr)
options(tigris_class = "sf")

install_github("jamgreen/lehdr")
library(lehdr)

#bring in lehd data for state and filter total emp and mfg
years <- 2004:2015

or_lehd <- grab_lodes(state = "or", year = years, lodes_type = "wac",
                      job_type = "JT01", segment = "S000", agg_geo = "bg")

or_lehd <- or_lehd %>% 
  select(1:4, CNS07)

names(or_lehd) <- tolower(names(or_lehd))
```

## Set Up (cont.)

```
#download state block group file
#create new county column

or_bg <- block_groups(state = "or") %>% 
  select(1:2,5, geom = geometry)

names(or_bg) <- tolower(names(or_bg))


#filter down to multco and join to the shapefile

or_lehd <- or_lehd %>% 
  mutate(county_fips = str_sub(w_bg, 1, 5)) %>% 
  filter(county_fips == "41051")
```

## Set Up (cont.)

```
or_lehd <- or_lehd %>% 
  left_join(or_bg, by = c("w_bg" = "geoid"))

#prep year into proper date and export

or_lehd <- or_lehd %>%
  mutate(lehd_date = paste0(year,"-01-01"))

or_lehd$lehd_date <- as.Date(or_lehd$lehd_date, 
format = "%Y-%m-%d")
```

## Export Shapefile

```
or_lehd <- or_lehd %>% 
  st_as_sf(sf_column_name = "geom")

or_lehd <- or_lehd %>% 
  st_transform(crs = 2992)

st_write(or_lehd, "multco_mfg_lehd.shp")
```

##  One result

# The Dissertation

## Problem Statement

Nancey Green-Leigh (2012) exposes smart growth's "blindisde" concerning the management of industrial land. A lot of "sustainable development" policies and approaches ignore or are explicitly hostile to industrial land in cities. 

More personal question: why would a city convert a healthy industrial district? PDX protects industrial land, does it work?

## Research Questions

>- Who is actually protecting industrial land? What are the features of these policies?
>- In cities that do protect industrial land, has it been effective?
>- How do planners view industrial land and the political/institutional contexts in which they operate? How does industrial land protection policy transform planning practice?

## Who's protecting industrial land?

- I reviewed the zoning codes and comp plans of the fifty largest cities in the US in order to identify protective policies
- Had two rules for identification
  - The policy must be distinct from already existing industrial zoning 
  - The language of the policy should frame industrial land as an *asset*

## An archteypal example

Chicago's planned manufacturing district ordinance mentions fostering the city's industrial base and diversifying the economy as primary goals.

## Quick Results

- I've identified 11 cities out of the 50 largest that have some form of industrial land policy
- Cities use a variety of approaches to enforcement largely under either additional "use" restrictions or more burdensome "process"

##  Do these policies work?

My most computationally intensive chapter. Basic question: do industrial land protections actually protect industrial jobs?

## Data

- LEHD WAC data for the 48 largest cities from 2004, 2009, 2015 at block group level
- 2006-2010 ACS data, housing prices, % renters, race/ethnicity at block group level
- City center points
- Spatial polygons for industrial districts

## Methodology- Data Preparation

- used *lehdr* and *tidycensus* to collect LEHD and ACS data, respectively
- had to create NYC IBZs from PDF and add to industrial district layer
- use PostGIS, from R, to get distance measures from city center to each block group
- create dummy variables for block groups where their centroids fell within industrial districts
- grab manufacturing employment from LEHD and "industrial" employment including warehousing/logistics, mining/ag, and construction

## Methodology- Analysis

- estimate inverse probability weighted propensity scores using *ipw* package and the ACS data - run a basic regression model with dependent variable $\Delta Mfg. Emp$, $\Delta Ind. Emp$ and indepedent variable $\beta_{1}District.dummy_{propensity score}$

## Results

>- Industrial district protection do not, in my specification, protect *manufacturing jobs*
>- Industrial district protection does, in my specification, protect a more widely defined class of *industrial* jobs
>-  Will continue to troubleshoot model

## How do planners view industrial land, in practice?

- Final chapter seeks to figure out how planners view industrial land preservation and what institutional changes, if any, accompany these new policies
- This matters because these policies signal a potential new hybrid economic development/land use strategy that challenges traditional planning roles and institutions

## Methodology

- Wrote a survey and sent to planning and economic development offices in the nation's fifty largest cities
- Sought professionals who deal directly with industrial land management/development
- Asked questions about personal views of the policy, any political pressure planners may receive around industrial land issues and city strategies
- Separately reviewed planning documents and organizational structure to try and suss out institutional change/transformation

## Results

- 18/50 responses. Not terrible response rate, not attempting to generalize. Purely descriptive.
- Respondents generally view industrial land as an *asset*, most projected cities can meet industrial land demand into medium term (10-15 years) but cities with protective policies largely expressed acute shortages (1-5 years)
- Industrial strategy is often divorced from traditional land-use purview. A potential inefficiency and gap. 
- Cities with industrial policies do exhibit some substantive institutional and practical change

## Results- Institutional Change

>-  New/additional reporting requirements (industrial land audits)
>-  Industrial land development reviews require new approaches or additional scrutiny
>-  Industrial land is now actively managed like any other city resource, requiring additional staff time
>-  Potentially new connections between development and planning agencies
>-  The rise of intermediaries responsible for industrial district support

