---
title: 'Census, LEHD and Planning: An R Approach'
author: "Jamaal Green"
institute: "Portland State University"
date: "12/10/2018"
output: 
  beamer_presentation:
    theme: "Montpellier"
    colortheme: "beetle"
    fonttheme: "structurebold"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# Introduction


## Good Morning

- Doctoral candidate here in TSUSP, planning on defending this spring
- My dissertation focuses on issues surrounding industrial land use policy
- My GRA work deals with a mix of mapping, some spatial analysis, census data wrangling and some modeling

## Goals for today

- Discuss *tidycensus* and my day-to-day workflows
- Introducing *lehdr*
- Dissertating with R

# Tidycensus and the Day-to-Day

## Tidycensus- Game Changer

- *tidycensus* is an R packaged authored and maintained by Prof. Kyle Walker at TCU
- a wrapper of the censusapi package and simplifies calls to grab ACS data, in particular
- radically simplifies and streamlines traditional census workflows, particularly with GIS and gives us data in a "tidy" format

## Tigris

- a package for grabbing Tiger/Line shapefiles
- also authored by Kyle Walker
- *tidycensus* depends on tigris bringing tabular and spatial data calls in one


## My old workflow

- Tabular Data collection/import (FactFinder or DataFerret for ACS/Decennial)
- Spatial Data collection(Tigerline files)
- Tabular data cleaning and munging in excel 
- Table to spatial data joins in Arc (successful maybe 40% of the time on first attempt) 
- If making maps...spatial processing (clips, intersections, spatial joins)
- Other visualizations and report writing (excel, word, and ppt)

##  Can tidycensus do all that?!

>-  No
>-  But tidycensus, in particular, directly solves aspects of the first three areas
>-  In combination with additional packages from the tidyverse (particularly *dplyr* and *ggplot*) and the spatial package *sf* all data processing can be done with R

## An Example

Our assignment: Get latest 5 year MHI for Multnomah County at Tract Level and graph the results (as an added bonus, and in the interest of transparency, let's include CVs)

```
install.packages("pacman"); library(pacman)
p_load(ggplot2, tidycensus, dplyr)

acs_key <- Sys.getenv("CENSUS_API_KEY")

#Enter the variables and geographies below
census_title <- c("Median Household Income by County:\n 
Margin of Error")
census_var <- c("B19013_001E")
census_geog <- c("county")
census_state <- c("or")
```

## The tidycensus call

```
acs_data <- get_acs(geography = census_geog, variables = 
census_var, state = census_state, output = "wide")

#Make more readable column names
acs_data <- acs_data %>% 
rename(MHI_est = B19013_001E , MHI_moe = B19013_001M)

#Calculate the SE, CV for future reference
acs_data <- acs_data %>% 
mutate(se = MHI_moe/1.645, cv = (se/MHI_est)*100)

```

## Finally, to the plot...

```
#Plot Percentages with Derived MOE
acs_plot <- acs_data %>%
  ggplot(aes(x = MHI_est, y = reorder(NAME, MHI_est))) +
  geom_point(color = "black", size = 2) +
  geom_errorbarh(aes(xmin = MHI_est - MHI_moe, 
  xmax = MHI_est + MHI_moe )) +
  labs(title = paste(census_title),
       subtitle = paste0("Oregon 2011-2015 American 
       Community Survey"),
       y = "",
       x = "Median Household Income") + 
  scale_x_continuous(labels = scales::dollar) + 
  theme_minimal() +
  theme(panel.grid.minor.x  = element_blank(),
        panel.grid.major.x = element_blank())

plot(acs_plot)
```


## Sample Output 

```{r echo=FALSE, message=FALSE, warning=FALSE}
if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(ggplot2, tidycensus, dplyr)


#key <- census_api_key("c67f1b3134300374d51a55c543649f843fb7d2b3",install = TRUE)
acs_key <- Sys.getenv("CENSUS_API_KEY")

#Enter the variables and geographies below
census_title <- c("Median Household Income by County:\n Margin of Error")
census_var <- c("B19013_001E")
census_geog <- c("county")
census_state <- c("or")


acs_data <- get_acs(geography = census_geog, variables = census_var, state = census_state, 
                    output = "wide")

#Make more readable column names
acs_data <- acs_data %>% rename(MHI_est = B19013_001E , MHI_moe = B19013_001M)

#Calculate the SE, CV for future reference
acs_data <- acs_data %>% mutate(se = MHI_moe/1.645, cv = (se/MHI_est)*100)



#Plot Percentages with Derived MOE
acs_plot <- acs_data %>%
  ggplot(aes(x = MHI_est, y = reorder(NAME, MHI_est))) +
  geom_point(color = "black", size = 2) +
  geom_errorbarh(aes(xmin = MHI_est - MHI_moe, xmax = MHI_est + MHI_moe )) +
  labs(title = paste(census_title),
       subtitle = paste0("Oregon 2011-2015 American Community Survey"),
       y = "",
       x = "Median Household Income") + 
  scale_x_continuous(labels = scales::dollar) + theme_minimal() +
  theme(panel.grid.minor.x  = element_blank(),
        panel.grid.major.x = element_blank())

print(acs_plot)
```

## Extending the package

 Let's make it spatial...going to map MHI at tract level for the PDX MSA
 
## Setting it up
```
mhi_tables <- c("B19013_001")

mhi_tract <- get_acs(geography = "tract", 
                     variables = mhi_tables, 
                     state = "OR", 
                     geometry = TRUE)

mhi_tract <- mhi_tract %>% 
  mutate(CountyFIPS = str_sub(GEOID, 1, 5))

metro_counties <- c("41051", "41005", "41009", "41067", 
"41071") 

metro_tract <- mhi_tract %>% 
  filter(CountyFIPS %in% metro_counties)
  
```
\end{small}

## Finalizing plot with *sf* and *ggplot*

```
p1 <- ggplot() + 
  geom_sf(data = metro_tract, aes(fill = estimate)) +
  coord_sf(datum = NA) + 
  theme(plot.title = element_text(size = 16, 
                    face = "bold", margin = margin(b=10))) +
  theme(plot.subtitle = element_text(size = 14, 
                      margin = margin(b = -20))) +
  theme(plot.caption = element_text(size = 9, 
                    margin = margin(t = -15), hjust = 0)) +
  scale_fill_viridis(labels = scales::dollar, 
                     name = "MHI Estimate") +
  labs(caption = "Source: US Census Bureau ACS (2011-2015)",
        title = "Median Household Income for PDX Metro\n 
        at the census tract level",
        subtitle = "An R 'sf' Example") + theme_minimal()


```

## Our result

```{r echo=FALSE, message=FALSE, warning=FALSE}

if(!require(pacman)){install.packages("pacman"); 
  library(pacman)}
p_load(sf, tigris, viridis, ggthemes, ggplot2, 
       tidycensus, stringr, dplyr)
options(tigris_class = "sf", tigris_use_cache = TRUE)

mhi_tables <- c("B19013_001")


mhi_tract <- get_acs(geography = "tract", 
                     variables = mhi_tables, 
                     state = "OR", 
                     geometry = TRUE)

mhi_tract <- mhi_tract %>% 
  mutate(CountyFIPS = str_sub(GEOID, 1, 5))

metro_counties <- c("41051", "41005", "41009", "41067", "41071") 
metro_tract <- mhi_tract %>% 
  filter(CountyFIPS %in% metro_counties)

p1 <- ggplot() + 
  geom_sf(data = metro_tract, aes(fill = estimate)) +
  coord_sf(datum = NA) + 
  theme(plot.title = element_text(size = 16, 
                    face = "bold", margin = margin(b=10))) +
  theme(plot.subtitle = element_text(size = 14, 
                      margin = margin(b = -20))) +
  theme(plot.caption = element_text(size = 9, 
                    margin = margin(t = -15), hjust = 0)) +
  scale_fill_viridis(labels = scales::dollar, 
                     name = "MHI Estimate") +
  labs(caption = "Source: US Census Bureau ACS (2011-2015)",
        title = "Median Household Income for PDX Metro\n at the census tract level",
        subtitle = "An R 'sf' Example") + theme_minimal()

plot(p1)

```

## From there

- From here I would use RMarkdown for report writing, back up with version control in github and go from there
- While a simplistic example, *tidycensus* along with other packages have freed my daily workflow from the time suck of going back and forth from FactFinder or DataFerret for common census datasets as well as switching from desktop GIS to excel

# *lehdr*- Freeing LODES data

## *lehdr*

- Package I designed with Profs. Liming Wang here at PSU and Dillon Mahmoudi at UMBC
- Comes out of my dissertation work to more efficienctly grad LEHD LODES data
- Hope to release on CRAN in 2019

## LEHD LODES

- Longitudinal Employer-Household Dynamics Dataset, Longitudinal Origin-Destination Employment Series
- A mouthful, but LODES gives us the finest geographic detail on employment (only 2 digit NAICS) for the country, as well as origin-destination tables
- You may be more familiar with [OnTheMap](https://onthemap.ces.census.gov/)

## Inspiration

- Dissertation requires looking at change in mfg employment between 2009-2015 at block group level
- Needed to grab multiple states for multiple time periods without using the ftp site manually
- Built up on from an unwiedly to relatively neat package designed to play nicely with tidy tools

## Manufacturing in Multnomah County

- My dissertation looks at urban mfg and land use
- Can we take a look at how the geography of mfg employment has changed over time in Multnomah County (2005-2015)

## The Set Up

```
if(!require(pacman)){install.packages("pacman"); 
library(pacman)}
p_load(devtools, dplyr, tigris, sf, stringr)
options(tigris_class = "sf")

install_github("jamgreen/lehdr")
library(lehdr)

#bring in lehd data for state and filter total emp and mfg
years <- 2004:2015

or_lehd <- grab_lodes(state = "or", year = years, lodes_type = "wac",
                      job_type = "JT01", segment = "S000", agg_geo = "bg")

or_lehd <- or_lehd %>% 
  select(1:4, CNS07)

names(or_lehd) <- tolower(names(or_lehd))
```

## Set Up (cont.)

```
#download state block group file
#create new county column

or_bg <- block_groups(state = "or") %>% 
  select(1:2,5, geom = geometry)

names(or_bg) <- tolower(names(or_bg))


#filter down to multco and join to the shapefile

or_lehd <- or_lehd %>% 
  mutate(county_fips = str_sub(w_bg, 1, 5)) %>% 
  filter(county_fips == "41051")
```

## Set Up (cont.)

```
or_lehd <- or_lehd %>% 
  left_join(or_bg, by = c("w_bg" = "geoid"))

#prep year into proper date and export

or_lehd <- or_lehd %>%
  mutate(lehd_date = paste0(year,"-01-01"))

or_lehd$lehd_date <- as.Date(or_lehd$lehd_date, 
format = "%Y-%m-%d")
```

## Export Shapefile

```
or_lehd <- or_lehd %>% 
  st_as_sf(sf_column_name = "geom")

or_lehd <- or_lehd %>% 
  st_transform(crs = 2992)

st_write(or_lehd, "multco_mfg_lehd.shp")
```
