---
title: 'Census, LEHD and Planning: An R Approach'
author: "Jamaal Green"
institute: "Portland State University"
date: "12/10/2018"
output: 
  beamer_presentation:
    theme: "Montpellier"
    colortheme: "beetle"
    fonttheme: "structurebold"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# Introduction


## Good Morning

- Doctoral candidate here in TSUSP, planning on defending this spring
- My dissertation focuses on issues surrounding industrial land use policy
- My GRA work deals with a mix of mapping, some spatial analysis, census data wrangling and some modeling

## Goals for today

- Discuss *tidycensus* and my day-to-day workflows
- Introducing *lehdr*
- Dissertating with R

# Tidycensus and the Day-to-Day

## Tidycensus- Game Changer

- *tidycensus* is an R packaged authored and maintained by Prof. Kyle Walker at TCU
- a wrapper of the censusapi package and simplifies calls to grab ACS data, in particular
- radically simplifies and streamlines traditional census workflows, particularly with GIS and gives us data in a "tidy" format

## Tigris

- a package for grabbing Tiger/Line shapefiles
- also authored by Kyle Walker
- *tidycensus* depends on tigris bringing tabular and spatial data calls in one


## My old workflow

- Tabular Data collection/import (FactFinder or DataFerret for ACS/Decennial)
- Spatial Data collection(Tigerline files)
- Tabular data cleaning and munging in excel 
- Table to spatial data joins in Arc (successful maybe 40% of the time on first attempt) 
- If making maps...spatial processing (clips, intersections, spatial joins)
- Other visualizations and report writing (excel, word, and ppt)

##  Can tidycensus do all that?!

>-  No
>-  But tidycensus, in particular, directly solves aspects of the first three areas
>-  In combination with additional packages from the tidyverse (particularly *dplyr* and *ggplot*) and the spatial package *sf* all data processing can be done with R

## An Example

Our assignment: Get latest 5 year MHI for Multnomah County at Tract Level and graph the results (as an added bonus, and in the interest of transparency, let's include CVs)

```
install.packages("pacman"); library(pacman)
p_load(ggplot2, tidycensus, dplyr)

acs_key <- Sys.getenv("CENSUS_API_KEY")

#Enter the variables and geographies below
census_title <- c("Median Household Income by County:\n 
Coefficient of Variation")
census_var <- c("B19013_001E")
census_geog <- c("county")
census_state <- c("or")
```

## The tidycensus call

```
acs_data <- get_acs(geography = census_geog, variables = 
census_var, state = census_state, output = "wide")

#Make more readable column names
acs_data <- acs_data %>% 
rename(MHI_est = B19013_001E , MHI_moe = B19013_001M)

#Calculate the SE, CV for future reference
acs_data <- acs_data %>% 
mutate(se = MHI_moe/1.645, cv = (se/MHI_est)*100)

```

## Finally, to the plot...

```
#Plot Percentages with Derived MOE
acs_plot <- acs_data %>%
  ggplot(aes(x = MHI_est, y = reorder(NAME, MHI_est))) +
  geom_point(color = "black", size = 2) +
  geom_errorbarh(aes(xmin = MHI_est - MHI_moe, 
  xmax = MHI_est + MHI_moe )) +
  labs(title = paste(census_title),
       subtitle = paste0("Oregon 2011-2015 American 
       Community Survey"),
       y = "",
       x = "Median Household Income") + 
  scale_x_continuous(labels = scales::dollar) + 
  theme_minimal() +
  theme(panel.grid.minor.x  = element_blank(),
        panel.grid.major.x = element_blank())

plot(acs_plot)
```


## Sample Output 

```{r echo=FALSE, message=FALSE, warning=FALSE}
if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(ggplot2, tidycensus, dplyr)


#key <- census_api_key("c67f1b3134300374d51a55c543649f843fb7d2b3",install = TRUE)
acs_key <- Sys.getenv("CENSUS_API_KEY")

#Enter the variables and geographies below
census_title <- c("Median Household Income by County:\n Coefficient of Variation")
census_var <- c("B19013_001E")
census_geog <- c("county")
census_state <- c("or")


acs_data <- get_acs(geography = census_geog, variables = census_var, state = census_state, 
                    output = "wide")

#Make more readable column names
acs_data <- acs_data %>% rename(MHI_est = B19013_001E , MHI_moe = B19013_001M)

#Calculate the SE, CV for future reference
acs_data <- acs_data %>% mutate(se = MHI_moe/1.645, cv = (se/MHI_est)*100)



#Plot Percentages with Derived MOE
acs_plot <- acs_data %>%
  ggplot(aes(x = MHI_est, y = reorder(NAME, MHI_est))) +
  geom_point(color = "black", size = 2) +
  geom_errorbarh(aes(xmin = MHI_est - MHI_moe, xmax = MHI_est + MHI_moe )) +
  labs(title = paste(census_title),
       subtitle = paste0("Oregon 2011-2015 American Community Survey"),
       y = "",
       x = "Median Household Income") + 
  scale_x_continuous(labels = scales::dollar) + theme_minimal() +
  theme(panel.grid.minor.x  = element_blank(),
        panel.grid.major.x = element_blank())

print(acs_plot)
```
