---
title: "Pragmatic Approach to Census Analysis"
author: "Jamaal Green"
date: "September 14, 2017"
output: 
  revealjs::revealjs_presentation:
    theme: "black"
    transition: slide
    self_contained: false
    reveal_plugins: ["zoom"]
    incremental: true
---

<script
  src="https://code.jquery.com/jquery-3.1.1.min.js"
  integrity="sha256-hVVnYaiADRTO2PzUGmuLJr8BLUSjGIZsDYGmIJLv2b8="
  crossorigin="anonymous"></script>

<script src="https://ajax.googleapis.com/ajax/libs/jqueryui/1.10.4/jquery-ui.min.js"></script>
<link href="https://ajax.googleapis.com/ajax/libs/jqueryui/1.10.4/themes/smoothness/jquery-ui.min.css" rel="stylesheet" type="text/css"/>

#Introduction
```{r, message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
```

##About Me

* PhD Candidate in Urban Studies and Planning
* My dissertation is examining industrial zoning and labor market change
* I use a pretty wide array of census products for work (ACS, PUMS, LEHD)

##A Pragmatic Approach

"Let the question guide your method"

Likewise...

Let your problems guide your tools

##What's your workflow normally resemble?

The steps many social data analysts and GIS user have to make:

* Tabular Data collection/import (for us clearly varied Census products)
* Spatial Data collection(Tigerline files, anyone?)
* Tabular data cleaning, munging, and joins
* Table to spatial data joins (we've all done this in Arc with moderate success)
* If making a maps...spatial processing (clips, intersections, spatial joins)
* Other visualizations and report writing

#Enter R

##What is R?

- A powerful language
- Application
- "Do it all" workbench

##But why R?

- It's free
- It's fast
- It's data type agnostic (read any variety of text files, .shp. GEOJSON, GEOTIFF)
- Massive number of packages for statistical or spatial analysis and visualization
- Many things that are hard or slow in other applications (table joins in Arc, anyone?) are fast in R 
- IT'S FREE

##But why should I?

Has the following ever happened to you?

- Need to download multiple variables over multiple years and you get a data folder filled with ambiguously named tables that you end up deleting anyway?
- Had to change your geography of interest on short notice and then go through the time consuming process of redownloading and processing?
- Attempted to rename a column in ArcMap (yes, I know this is now available in ArcPro)?

##Let's Be Pragmatic

<section style="text-align: left;">

These recurring challenges can be fully, or partially addressed, by learning a little bit of R 

Let's take an example...

</section>

#Tidycensus...A Better Way

##Tidycensus...one stop shop for ACS data

- R package authored by Prof. Kyle Walker at TCU to make gathering and visualizing census data easier
- The package uses the census API to call ACS and decennial data as well as ACS Data Profile tables
- Data is returned in either wide or long format and there is an option to join the data to its appropriate Tigerline geometry

##A quick example...

Our assignment: Get latest 5 year MHI for Multnomah County at Tract Level and graph the results (as an added bonus, and in the interest of transparency, let's include CVs)

```
if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(ggplot2, tidycensus, dplyr)

acs_key <- Sys.getenv("CENSUS_API_KEY")

#Enter the variables and geographies below
census_title <- c("Median Household Income by County:\n 
Coefficient of Variation")
census_var <- c("B19013_001E")
census_geog <- c("county")
census_state <- c("or")


acs_data <- get_acs(geography = census_geog, variables = 
census_var, state = census_state, output = "wide")

#Make more readable column names
acs_data <- acs_data %>% rename(MHI_est = B19013_001E , MHI_moe = B19013_001M)

#Calculate the SE, CV for future reference
acs_data <- acs_data %>% mutate(se = MHI_moe/1.645, cv = (se/MHI_est)*100)



#Plot Percentages with Derived MOE
acs_plot <- acs_data %>%
  ggplot(aes(x = MHI_est, y = reorder(NAME, MHI_est))) +
  geom_point(color = "black", size = 2) +
  geom_errorbarh(aes(xmin = MHI_est - MHI_moe, xmax = MHI_est + MHI_moe )) +
  labs(title = paste(census_title),
       subtitle = paste0("Oregon 2011-2015 American Community Survey"),
       y = "",
       x = "Median Household Income") + 
  scale_x_continuous(labels = scales::dollar) + theme_minimal() +
  theme(panel.grid.minor.x  = element_blank(),
        panel.grid.major.x = element_blank())

plot(acs_plot)

```

##Our Output

```{r, echo=FALSE, message=FALSE, warning=FALSE}
if(!require(pacman)){install.packages("pacman"); library(pacman)}
p_load(ggplot2, tidycensus, dplyr)


#key <- census_api_key("c67f1b3134300374d51a55c543649f843fb7d2b3",install = TRUE)
acs_key <- Sys.getenv("CENSUS_API_KEY")

#Enter the variables and geographies below
census_title <- c("Median Household Income by County:\n Coefficient of Variation")
census_var <- c("B19013_001E")
census_geog <- c("county")
census_state <- c("or")


acs_data <- get_acs(geography = census_geog, variables = census_var, state = census_state, 
                    output = "wide")

#Make more readable column names
acs_data <- acs_data %>% rename(MHI_est = B19013_001E , MHI_moe = B19013_001M)

#Calculate the SE, CV for future reference
acs_data <- acs_data %>% mutate(se = MHI_moe/1.645, cv = (se/MHI_est)*100)



#Plot Percentages with Derived MOE
acs_plot <- acs_data %>%
  ggplot(aes(x = MHI_est, y = reorder(NAME, MHI_est))) +
  geom_point(color = "black", size = 2) +
  geom_errorbarh(aes(xmin = MHI_est - MHI_moe, xmax = MHI_est + MHI_moe )) +
  labs(title = paste(census_title),
       subtitle = paste0("Oregon 2011-2015 American Community Survey"),
       y = "",
       x = "Median Household Income") + 
  scale_x_continuous(labels = scales::dollar) + theme_minimal() +
  theme(panel.grid.minor.x  = element_blank(),
        panel.grid.major.x = element_blank())

print(acs_plot)
```

##Let's Stretch a Bit

Let's say we now need to 